The 2013 SMT Evaluation

### 2015-02-16 2013/report/SMTEVAL-2013.tex
Moved (non-public) source files out of smtcomp-web.

\documentclass[smallcondensed]{svjour3}
%\providecommand{\event}{Journal TBD}
%\providecommand{\publicationstatus}{ } %% TBD - fill in any footer
%%\documentclass[12pt,oneside]{llncs}

%% Language %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{hyperref}   %% sets hyperlinks within a pdf
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{array}
\usepackage{color}
\usepackage{colortab}
\usepackage{xcolor}
\usepackage{tabularx}
\usepackage{pdfpages}

@@

%\def\titlerunning{SMTEVAL-2013}
%\def\authorrunning{Cok, Stump, Weber}

@@

\newcommand{\cellbr}[2][c]{\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\journalname{Journal of Automated Reasoning}

\begin{document}
\title{The 2013 Evaluation of SMT-COMP and SMT-LIB}

@@

%% eptcs format
\author{
David R. Cok 
\and  
Aaron Stump
\and  
Tjark Weber
}

\institute{
David R. Cok  \at GrammaTech, Inc., USA \email{dcok@grammatech.com}
\and
Aaron Stump \at University of Iowa, IA, USA \email{aaron-stump@uiowa.edu}
\and
Tjark Weber \at Uppsala University, Sweden \email{tjark.weber@it.uu.se}
}
%% lncls format
%\author{David R. Cok \inst{1} \and \\ 
%Aaron Stump \inst{2} \and \\ 
%Tjark Weber \inst{3}  \\
%}
%
%\institute{GrammaTech, Inc., USA \email{dcok@grammatech.com} \and
%University of Iowa, IA, USA \email{aaron-stump@uiowa.edu} \and
%Uppsala University, Sweden, \email{tjark.weber@it.uu.se}
%}

\date{}
\maketitle
%\pagestyle{headings}
%\pagenumbering{arabic}
%\centerline{\today}

@@

\subsection{The Competition history and goals}
From 2005 through 2012 (and in 2014), the SMT community
sponsored an annual competition among SMT solvers
(cf.\ Fig.~\ref{Fig:history}).  The purpose of the competition is to
encourage advances in SMT solver implementations acting on benchmark
formulas of theoretical or practical interest.  Public competitions
are a well-known means of stimulating advancement in software tools.
For example, in automated reasoning, the SAT and CASC competitions for
propositional and first-order reasoning tools, respectively, have
spurred significant innovation in their fields~\cite{leberre+03,PSS02}.  Indeed, the SMT competition increased in
size each year: more benchmarks were added, new solver teams
participated, solver performance improved, and more logic divisions were defined.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{SMTCOMP-History}
\caption{SMT-COMP history}
\label{Fig:history}
\end{figure}

@@

And third, for many years the competition had relied on the SMT-Exec computational infrastructure~\cite{springerlink:10.1007/s10817-012-9246-5}.  A new infrastructure, StarExec~\cite{StuST-IJCAR-14,webStarExec} had been funded by the NSF and developed at the University of Iowa, but had not yet been tried out in earnest.  In fact, a goal in~2012 had been to use StarExec for SMT-COMP~2012; however, StarExec was not sufficiently operational and the competition had to revert to SMT-Exec.  It was anticipated that an evaluation in 2013, with less deadline pressure, would enable using StarExec on a shake-down cruise prior to a competition in 2014 at the Federated Logic Conference~(FLoC) Olympic Games~\cite{webOlympicGames}.

@@

The SMT Steering Committee appointed a team of evaluators (the authors
of this report) to examine areas of interest. The evaluators, who are
not affiliated with any solver team, independently studied the
following topics:
\begin{itemize}[noitemsep]
\item continuity and turnover in SMT solver participation in past competitions (\ref{Solvers})
\item the performance of all historical and current solvers on the full set of benchmarks, measuring
\begin{itemize}[noitemsep]
\item the improvement in performance over time (\ref{Progress})
\item repeatability of performance measurements (\ref{Accuracy})
\item repeatability of competition results (\ref{CompetitionRepeatability})
\item competitiveness of solvers (\ref{Competitiveness})
\item parallel vs. sequential performance (\ref{parallel})
\end{itemize}
\item the usefulness of various logics
\begin{itemize}[noitemsep]
\item characteristics of existing logics (\ref{Lattice})
\item which logics are implemented by solvers (\ref{Coverage})
\item which logics are particularly relevant to application areas (\ref{Applications})
\end{itemize}
\item the state of existing benchmarks
\begin{itemize}[noitemsep]
\item the range of computational difficulty of the benchmarks (\ref{Benchmarks})
\item the degree to which benchmarks discriminate among solvers (\ref{Benchmarks})
\item which logics have support in benchmarks (\ref{Benchmarks})
\end{itemize}
\end{itemize}

@@

One goal of the SMT project since its inception has been collecting benchmarks by which to evaluate SMT solvers and to represent challenge problems in the field. The growth in the number of benchmarks available is shown in the graphics of Fig.~\ref{Fig:history}. All\footnote{One benchmark, added late, was not included in all of the data accumulation; hence some tables report 95\,491 benchmarks and some 95\,492.} of the benchmarks present in StarExec at the time of evaluation (95\,492 benchmarks) were used for the evaluation, including those with unknown status. Of course, in previous years only a subset of these were available, since new benchmarks were added from year to year. Also in 2012, some benchmarks were used in the competition that were not yet added into the SMT-LIB benchmark set and were not yet in StarExec at the time of the evaluation. The 2012 benchmark set also included some incremental benchmarks that were not used in this evaluation. Furthermore, the competition each year used a random selection of benchmarks (guided by a difficulty distribution). The fact that the set of benchmarks used in competitions was different each year muddied any year-to-year comparison of results. By using all available benchmarks in the evaluation, we ensure that the current assessment is not susceptible to random choices in benchmark selection (cf.~Section~\ref{CompetitionRepeatability}). The various kinds of benchmarks and their distribution across logics are discussed below in Section~\ref{Evaluations}.

\subsection{Solvers}
\label{sec:solvers}

@@

\begin{figure}
\centering
\begin{tabular}{|l|l|*{8}{c@{\hskip 10pt}}c|}
\hline
\input{solvers2}
\end{tabular}

\caption{Solvers used in each year of SMT-COMP and SMT-EVAL. Solvers prior to 2010 (marked by~X) do not support SMT-LIB~v2 and were not used for SMT-EVAL. Boxes marked with a~\textbullet{} symbol identify solvers that are new for the evaluation; all solvers since 2010 (marked with~+ or~\textbullet{} symbols) were used for the evaluation.}
\label{Fig:solvers}
\end{figure}

@@

Given the large number of solver-benchmark combinations, it was not
feasible to manually inspect the output of each job pair.  Thus, we
needed to determine mechanically whether a solver reported a benchmark
as satisfiable, unsatisfiable or unknown (or did not return a result
within the time limit).

On StarExec, this is done by providing a \emph{post-processor}, i.e.,
an executable that operates on the output of each job pair.
Unfortunately, two features of StarExec made it difficult to interpret
solver output in a way that is fully SMT-LIB~v2 compliant.  First,
StarExec collects the entire output of a job pair in a single file
before passing it to the post-processor.  Thus, it is not possible to
reliably determine which output a solver has generated specifically in
response to a benchmark's {\tt check-sat} command.  Second, StarExec
conflates standard output and standard error.  We hope that this
behavior will be changed in the future, but for SMT-EVAL this made it
impossible to distinguish between regular solver responses and error
messages.

To determine the result of each job pair, we used a post-processor
that searched for the words {\tt sat}, {\tt unsat} and {\tt unknown}
in the output file, and reported the corresponding result if it found
exactly one of them; otherwise, it reported no result.  This
conservative approach may have caused a small number of (otherwise
valid) solver responses to be discarded.  It was adopted to minimize the
number of incorrect results reported by the post-processor.

@@

An additional use of benchmark scrambling is that the change in syntax
and identifiers can, for some solvers, change the search paths or the
preprocessing that is performed. Thus scrambling serves to exercise
different code paths and adds an additional dimension of testing and
performance measurement to competitions and evaluations. It would be
interesting to investigate these effects of scrambling in detail; we
did not do this as part of SMT-EVAL.

%Several solvers reported syntax errors on some benchmarks.  We
%validated all benchmarks with two independently developed SMT-LIB~v2
%parsers~\cite{TODO,TODO} to ensure that they are syntactically
%conforming.

\subsection{StarExec}

SMT-EVAL successfully used the new StarExec computational framework~\cite{StuST-IJCAR-14,webStarExec}. Running SMT-EVAL on StarExec did indeed expose a number of bugs and user interface issues; these were corrected in the course of the SMT-EVAL runs. Thus SMT-EVAL served a valuable purpose in preparing StarExec for larger scale use.

SMT-EVAL's largest computational job was running all 45 of the evaluated solvers on all of the relevant non-incremental benchmarks in the SMT library. 
The benchmarks belong to different {\em logics} and solvers are characterized by which logics they support. 
So for each logic, the evaluation executed the cross product of all benchmarks for that logic and all solvers (in all years) that support problems in that logic, for a total of 1\,663\,478 solver-benchmark combinations (called {\em job-pairs} in StarExec). 
This job took several months of wall-clock time to run. 
We ran it in quarters, using the result of the first quarter to adjust our procedures and debug some of StarExec, before running the rest of the solver-benchmark job pairs. 
Because of a bug (now corrected), the results of about 600 job-pairs were not present in the accumulated results. 
These were identified and rerun as an additional ``mop-up'' job. 
Each solver-benchmark pair was run in an independent environment, on a computation node by itself, with time and memory limits, as described in Section~\ref{parallel}.

@@

The evaluation team's observations on the questions it considered are presented in the following subsections; our overall conclusions are listed in Section~\ref{Conclusions}.  The raw data used as the basis for our observations, collected from StarExec, are all archived on the SMT-COMP website, at \url{http://sourceforge.net/p/smtcomp/code/HEAD/tree/trunk/smtcomp-web/2013/data} as 7z-compressed files.  (The largest is~25\,MB.)  The data can also be viewed directly on StarExec, at \url{https://www.starexec.org/starexec/secure/explore/spaces.jsp?id=4566}, using its guest login.

@@

\begin{figure}
\centering
\begin{tabular}{|l|ccccccccc|}
\hline
year                & 2005 & 2006 & 2007 & 2008 & 2009 & 2010 & 2011 & 2012 & 2013 \\ \hline
\# of participants  & 11   & 11   & 9    & 13   & 12   & 10   & 11   & 11   & 9   \\ \hline
\# dropping out     &      &  4   & 6    &  2   &  3   &  7   &  2   &  4   & \\ \hline
\# new participants & (11) &  4   & 4    &  6   &  2   &  5   &  3   &  4   &      \\ \hline
\end{tabular}
\caption{Turnover in solver team participation. There are no entries for changes in 2013 because in that year there was no competition with entrants.}
\label{Fig:participation}
\end{figure}
\begin{itemize}
\item As shown in Fig. \ref{Fig:participation}, the number of solvers participating each year has been fairly constant, ranging
from 9-13, with a median value of 11 participants. The data for 2013 is an anomaly in two respects. The count of 9 only includes those solvers that were explicitly submitted new for 2013. Second, Abziz contributed a portfolio solver that makes an automatic choice, based on machine learning, of which among other existing solvers to apply to a benchmark using observed features of the benchmark. Abziz submitted two instances of his portfolio that used solvers from 2011; those instances of his portfolio solver are not counted in 
Fig. \ref{Fig:participation} since they were not available in 2011, though they are included in Fig.~\ref{Fig:solvers}. In addition, three variations were contributed in 2012; we count just one participation unit in 
Fig. \ref{Fig:participation} for 2012, though all three (and the two 2011 entries) were used in the evaluations described in this paper.

@@

\subsection{Logics}
\label{Lattice}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{LogicLatticeN}
\caption{Graph of SMT logics defined in SMT-LIB. The links represent a natural inclusion by characteristics, but {\it not} a sub-logic relationship. QF\_ABV is defined, but is often subsumed in QF\_AUFBV, and is not usually listed separately in other tables in this report. Q is not a logic by itself but is a characteristic indicating that quantified expressions are permitted in the logic; similarly N indicates non-linear arithmetic is allowed.}
\label{Fig:lattice}
\end{figure}

SMT-LIB defines a number of theories and logics. The logics are a combination of theories with additional functions or logic symbols or restrictions on the vocabulary of the theories. For example, the QF\_IDL logic uses the underlying Ints theory, but restricts terms to equalities and inequalities between constants and simple differences of variables.
The combination of the Ints theory and the Reals theory adds functions that convert between integers and reals (among other things).

Leaving aside difference logics as special cases of integer and real logics, there are essentially these mostly orthogonal characteristics underlying the theories and logics:
\begin{itemize}[noitemsep,nolistsep] %% these modifiers are not supported in all document classes
\item IA: integers 
\item RA: reals
\item N: non-linear arithmetic
\item A or AX: arrays
\item UF: uninterpreted functions
\item QF\_: quantifier freeness
\item BV: bit-vectors
\end{itemize}
This list of characteristics partially corresponds to the set of SMT-LIB theories. The exceptions are quantification and non-linearity, both of which designate lifting restrictions on the kinds of terms allowed in the resulting logic. (Since they do not correspond exactly to underlying SMT theories, the term {\em characteristics} is used in this discussion.)

It is possible to define a logic with any powerset of these characteristics in an almost purely combinatorial sense. 
Barring non-linearity without arithmetic and the useless empty set, there are 111 combinations of characteristics (instead of the full 128). SMT-LIB has definitions and benchmarks for 20 of these logics (excluding QF\_IDL, QF\_RDL, and QF\_UFIDL) at the time of the evaluation, expanded to 34 in 2014. Fig.~\ref{Fig:lattice} shows what might be expected to be natural containment relationships among the logics defined at the time of this evaluation.

However, what might be expected to be natural containment relationships are not actually sub-logic relationships. That more accurate relationship is shown in a diagram on the SMT-LIB website: \url{http://smtlib.cs.uiowa.edu/logics.shtml} (due to Cesare Tinelli). The principal cause of this lack of sub-logic relationship is that many of the logics supporting arrays support only specific sorts of arrays. For example, the basic QF\_AX logic allows arbitrary new sort symbols, including arrays over those free sorts. However, QF\_ABV (arrays and bit-vectors) does not allow new sort symbols and only allows arrays over bit-vector sorts.

The naming convention for logics is almost but not quite simply a combination of the letters indicated in the list above,
corresponding to the characteristics of the logic, with the letters listed in order by convention:
\begin{center}
[QF\_][A\textbar AX][UF][BV][N\textbar L][IA\textbar RA\textbar IRA].
\end{center}

\noindent The non-uniformities are these:
\begin{itemize}[noitemsep,nolistsep]
\item quantification is indicated by {\em removing} a QF\_ prefix;
\item the absence of non-linearity~(N), i.e., linearity, is indicated by an~L, instead of by no designator;
\item integer and real arithmetic are indicated by two letters (IA and~RA) and their combination by~IRA;
\item the~A used in the arithmetic designators could be ambiguous with the designator for arrays, except for position;
\item a logic with just arrays is named QF\_AX, whereas in other combinations including arrays, just an~A instead of~AX is used.
\end{itemize}
For example, UFNIA includes quantification (no QF\_ prefix), uninterpreted functions~(UF), non-linearity~(N), and integer arithmetic~(IA), leaving out arrays, bit-vectors, and reals.

These names have roots in the history of development of SMT-LIB. Nevertheless as the set of logics grows---23 here and 34 by the 2014 competition---the situation will increasingly confuse new users of SMT-LIB trying to understand the structure of logics; in addition new theories and accompanying logics are anticipated.
It would be useful to standardize the naming convention along the lines of current use, perhaps with slight adjustments (e.g., renaming QF\_AX to QF\_A and removing the excess~A from~IA and~RA), in order to define precisely the logic implied by any combination of designators. A policy regarding names would provide a process for defining names for new theories and logics. This would
also make it easier to accurately characterize new benchmarks and to summarize the logics supported by a given solver. This need has been raised to the SMT-LIB coordinators and some work has been started to address it.

@@

\begin{figure}
\centering
\input{LogicSupportBySolvers}
\caption{Support for each logic as stated by solver implementors. The dots instead of Xs identify the 2013 versions of solvers. The numbers in the bottom row are the number of 2013 solvers for each logic.}
\label{Fig:solversupport}
\end{figure}

@@

Note that the support indicated is that stated by the solver supplier at the time the solver was submitted to a competition. 
In some cases, a solver supports one logic but is not listed as supporting a subset of that logic. The general reason is that the subset, with restricted characteristics, allows for specific optimizations that are not implemented. The more general solver could act on problems of the
more restricted benchmarks but is not deemed competitive and thus was not formally entered into the competition for the restricted logic.

@@

Fig.~\ref{Fig:CumulativeTimes} shows the fraction of benchmarks completed (y-axis) within a given time (in seconds, on the x-axis); thus points toward the upper left are better (more completed in less time). The results for all logics are shown together. 
The four curves are the data for the four years from 2010 to 2013. The lowest (marked with fine dots) curve is that for 2010. 
There is clearly significant progress made from 2010 to later years: the number of benchmarks completed within a given time is noticeably higher; we did not attempt to identify the source of this historical improvement. The curves for 2011 and 2012 (heavy dashes and short dashes, respectively) are clearly above 2010, but are not substantially different from each other. The solvers for 2013 (heavy solid line) are uniformly above those for previous years, but not by a large amount.

@@

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{CumulativeTimes}
\caption{The fraction of benchmarks (y-axis) whose winning time is less than the given number of seconds (x-axis), by year.}
\label{Fig:CumulativeTimes}
\end{figure}

@@

%\section*{Acknowledgments} 
\begin{acknowledgements}

>>>>

\documentclass{eptcs}
\providecommand{\event}{Journal TBD}
\providecommand{\publicationstatus}{ } %% TBD - fill in any footer
%%\documentclass[12pt,oneside]{llncs}

%% Language %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{hyperref}   %% sets hyperlinks within a pdf
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{array}
\usepackage{color}
\usepackage{colortab}
\usepackage{xcolor}
\usepackage{tabularx}

@@

\def\titlerunning{SMTEVAL-2013}
\def\authorrunning{Cok, Stump, Weber}

@@

\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.8}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\floatpagefraction}{0.75}

\begin{document}

\title{The 2013 SMT Evaluation}

@@

%% eptcs format
\author{
David R. Cok \institute{GrammaTech, Inc., USA \email{dcok@grammatech.com}}
\and
Aaron Stump \institute{University of Iowa, IA, USA \email{aaron-stump@uiowa.edu}}
\and
Tjark Weber \institute{Uppsala University, Sweden \email{tjark.weber@it.uu.se}}
}

%% lncls format
%\author{David R. Cok \inst{1} \and \\
%Aaron Stump \inst{2} \and \\
%Tjark Weber \inst{3}  \\
%}
%
%\institute{GrammaTech, Inc., USA \email{dcok@grammatech.com} \and
%University of Iowa, IA, USA \email{aaron-stump@uiowa.edu} \and
%Uppsala University, Sweden, \email{tjark.weber@it.uu.se}
%}

\maketitle
\pagestyle{headings}
\pagenumbering{arabic}
\centerline{\today}

@@

\subsection{The Competition history and goals}
From 2005 through 2012 (and coming again in 2014), the SMT community
sponsored an annual competition among SMT solvers
(cf.\ Fig.~\ref{Fig:history}).  The purpose of the competition is to
encourage advances in SMT solver implementations acting on benchmark
formulas of theoretical or practical interest.  Public competitions
are a well-known means of stimulating advancement in software tools.
For example, in automated reasoning, the SAT and CASC competitions for
propositional and first-order reasoning tools, respectively, have
spurred significant innovation in their fields
\cite{leberre+03,PSS02}.  Indeed, the SMT competition increased in
size each year: more benchmarks were added, new solver teams
participated, solver performance improved, and more logic divisions were defined.

@@

\begin{figure}
\centering
\includegraphics[width=.95\textwidth, viewport=0 85 720 370, clip=true]{SMTCOMP-History}
\caption{SMT-COMP history}
\label{Fig:history}
\end{figure}

@@

And third, for many years the competition had relied on the SMT-Exec computational infrastructure\cite{springerlink:10.1007/s10817-012-9246-5}.  A new infrastructure, Star-Exec \cite{StuST-IJCAR-14,webStarExec} had been funded by the NSF and developed at the University of Iowa, but had not yet been tried out in earnest.  In fact, a goal in~2012 had been to use Star-Exec for SMT-COMP~2012; however, Star-Exec was not sufficiently operational and the competition had to revert to SMT-Exec.  It was anticipated that an evaluation in 2013, with less deadline pressure, would enable using Star-Exec on a shake-down cruise prior to a competition in 2014 at the Federated Logic Conference~(FLoC) Olympic Games \cite{webOlympicGames}.

@@

The SMT Steering Committee appointed a team of evaluators (the authors
of this report) to examine areas of interest. The evaluators studied
the following topics:
\begin{itemize}[noitemsep]
\item continuity and turnover in SMT solver participation in past competitions (\ref{Solvers})
\item the performance of all historical and current solvers on the full set of benchmarks, measuring
\begin{itemize}[noitemsep]
\item the improvement in performance over time (\ref{Progress})
\item repeatability of performance measurements (\ref{Accuracy})
\item repeatability of competition results (\ref{CompetitionRepeatability})
\item competitiveness of solvers (\ref{Competitiveness})
\item parallel vs. sequential performance (\ref{parallel})
\end{itemize}
\item the usefulness of various logics
\begin{itemize}[noitemsep]
\item characteristics of existing logics (\ref{Lattice})
\item which logics are implemented by solvers (\ref{Coverage})
\item which logics are particularly relevant to application areas (\ref{Applications})
\end{itemize}
\item the state of existing benchmarks
\begin{itemize}[noitemsep]
\item the range of computational difficulty of the benchmarks (\ref{Benchmarks})
\item the degree to which benchmarks discriminate among solvers (\ref{Benchmarks})
\item which logics have support in benchmarks (\ref{Benchmarks})
\end{itemize}

@@

One goal of the SMT project since its inception has been collecting benchmarks by which to evaluate SMT solvers and to represent challenge problems in the field. The growth in the number of benchmarks available is shown in the graphics of Fig.~\ref{Fig:history}. All\footnote{One benchmark, added late, was not included in all of the data accumulation; hence some tables report 95\,491 benchmarks and some 95\,492.} of the benchmarks present in Star-Exec at the time of evaluation (95\,492 benchmarks) were used for the evaluation, including those with unknown status. Of course, in previous years only a subset of these were available, since new benchmarks were added from year to year. Also in 2012, some benchmarks were used in the competition that were not yet added into the SMT-LIB benchmark set and were not yet in StarExec at the time of the evaluation. The 2012 benchmark set also included some incremental benchmarks that were not used in this evaluation. Furthermore, the competition each year used a random selection of benchmarks (guided by a difficulty distribution). The fact that the set of benchmarks used in competition was different each year muddied any year-to-year comparison of results. By using all available benchmarks in the evaluation, we intend that any comparative assessments across years or solvers be more accurate. The various kinds of benchmarks and their distribution across logics are discussed below in Section~\ref{Evaluations}.

\subsection{Solvers}

@@

\begin{figure}
\centering
\begin{tabular}{|l|l|c@{\hspace{-.1em}}ccccccccc|}
\hline
\input{solvers2}
\end{tabular}

\caption{Solvers used in each year of SMT-COMP and SMT-EVAL. Solvers prior to 2010 (marked by X) do not support SMT-LIB~v2 and were not used for SMT-EVAL.  Entries marked with a~\textbullet{} symbol identify solvers that are new for the evaluation.  All solvers since 2010 (marked with~+ or~\textbullet{} symbols) were used for the evaluation.}
\label{Fig:solvers}
\end{figure}

@@

@@

An additional use of benchmark scrambling is that the change in syntax and identifiers can, for some solvers,
change the search paths or the preprocessing that is performed. Thus scrambling serves to exercise
different code paths and adds an additional dimension of testing and performance measurement to
competitions and evaluations.

%Several solvers reported syntax errors on some benchmarks.  We
%validated all benchmarks with two independently developed SMT-LIB~v2
%parsers~\cite{TODO,TODO} to ensure that they are syntactically
%conforming.
\subsection{Star-Exec}

SMT-EVAL successfully used the new Star-Exec computational framework \cite{webStarExec,StuST-IJCAR-14}. Running SMT-EVAL on Star-Exec did indeed expose a number of bugs and user interface issues; these were corrected in the course of the SMT-EVAL runs. Thus SMT-EVAL served a valuable purpose in preparing Star-Exec for larger scale use.

SMT-EVAL's largest computational job was running all 45 of the evaluated solvers on all of the non-incremental benchmarks in the SMT library. The benchmarks belong to different {\em logics} and solvers are characterized by which logics they support. For each logic, the evaluation executed the cross product of all benchmarks for that logic and all solvers (in all years) that support problems in that logic, for a total of 1\,663\,478 solver-benchmark combinations (called {\em job-pairs} in Star-Exec). This job took several months of wall time to run. We ran it in quarters, using the result of the first quarter to adjust our procedures and debug some of Star-Exec, before running the rest of the solver-benchmark job-pairs. Because of a bug (now corrected), the results of about 600 job-pairs were not present in the accumulated results. These were identified and rerun as an additional ``mop-up'' job. Each job-pair was run in an independent environment, on a computation node by itself. Each node has four cores. The job-pair was limited to 25 minutes of wall-clock time, 100 minutes of cpu time, and 1GB of memory.

Given the large number of solver-benchmark combinations, it was not
feasible to manually inspect the output of each job-pair. Thus, we
needed to determine mechanically whether a solver reported a benchmark
as satisfiable, unsatisfiable or unknown (or did not return a result
within the time limit).

On StarExec, this is done by providing a \emph{post-processor}, i.e.,
an executable that operates on the output of each job-pair.
Unfortunately, two features of StarExec made it difficult to interpret
solver output in a way that is fully SMT-LIB~v2 compliant.  First,
StarExec collects the entire output of a job-pair in a single file
before passing it to the post-processor.  Thus, it is not possible to
reliably determine which output a solver has generated specifically in
response to a benchmark's {\tt check-sat} command.  Second, StarExec
conflates standard output and standard error, making it impossible to
distinguish between regular solver responses and error messages.

To determine the result of each job-pair, we used a post-processor
that searched for the words {\tt sat}, {\tt unsat} and {\tt unknown}
in the output file, and reported the corresponding result if it found
exactly one of them; otherwise, it reported no result.  This
conservative approach may have caused a small number of (otherwise
valid) solver responses to be discarded.  It was adopted to minimize the
number of incorrect results reported by the post-processor.

@@

The evaluation team's observations on the questions it considered are presented in the following subsections; our overall conclusions are listed in Section~\ref{Conclusions}.  The raw data used as the basis for our observations, collected from Star-Exec, are all archived on the SMT-COMP website, at \url{http://sourceforge.net/p/smtcomp/code/HEAD/tree/trunk/smtcomp-web/2013/data} as 7z-compressed files.  (The largest is~25\,MB.)

@@

\begin{figure}
\centering
\begin{tabular}{|l|cccccccc|c|}
\hline
Year                & 2005 & 2006 & 2007 & 2008 & 2009 & 2010 & 2011 & 2012 & 2013 \\ \hline
\# of participants  & 11   & 11   & 9    & 13   & 12   & 10   & 11   & 11   & 9 \\
\# dropping out     &      &  4   & 6    &  2   &  3   &  7   &  2   &  4   & \\
\# new participants & (11) &  4   & 4    &  6   &  2   &  5   &  3   &  4   & \\ \hline
\end{tabular}
\caption{Turnover in solver team participation. There are no entries for changes in 2013 because in that year there was no competition with entrants.}
\label{Fig:participation}
\end{figure}

\begin{itemize}
\item As shown in Fig. \ref{Fig:participation}, the number of solvers participating each year has been fairly constant, ranging
from 9-13, with a median value of 11 participants. The data for 2013 is an anomaly in two respects. The count of 9 only includes those solvers that were explicitly submitted new for 2013; some teams were glad to be included but did not upgrade their 2012 solvers. Second, Abziz contributed a portfolio solver that makes an automatic choice, based on machine learning, of which among other existing solvers to apply to a benchmark using observed features of the benchmark. Abziz submitted two instances of his portfolio that used solvers from 2011; those instances of his portfolio solver are not counted in
Fig. \ref{Fig:participation} since they were not available in 2011, though they are included in Fig.~\ref{Fig:solvers}. In addition, three variations were contributed in 2012; we count just one participation unit in
Fig. \ref{Fig:participation} for 2012, though all three (and the two 2011 entries) were used in the evaluations described in this paper.

@@

\subsection{Lattice of logics}
\label{Lattice}

\begin{figure}
\centering
\includegraphics[width=\textwidth,viewport=0 50 720 520, clip=true]{LogicLattice}
\caption{Lattice of SMT logics defined in SMT-LIB. The links represent a natural inclusion by characteristics, but {\it not} a sub-logic relationship. {\tt QF\_ABV} is defined, but is often subsumed in {\tt QF\_AUFBV}, and is not usually listed separately in other tables in this report. {\tt Q} is not a logic by itself but is a characteristic indicating that quantified expressions are permitted in the logic.}
\label{Fig:lattice}
\end{figure}

SMT-LIB defines a number of theories and logics. The logics are a combination of theories with additional functions or logic symbols or restrictions on the vocabulary of the theories. For example, the {\tt QF\_IDL} logic uses the underlying {\tt Ints} theory, but restricts terms to equalities and inequalities between constants and simple differences of variables.
The combination of the {\tt Ints} theory and the {\tt Reals} theory adds functions that convert between integers and reals (among other things).
Note that the {\tt QF\_ABV} logic is defined, but is often subsumed in {\tt QF\_AUFBV} and has no benchmarks of its own;\footnote{By the time of the 2014 competition, the benchmarks were reorganized and recategorized, and more benchmarks were added. So now this logic is populated as well.} it is not usually listed separately in other tables in this report.

Leaving aside difference logics as special cases of integer and real logics, there are essentially these mostly orthogonal characteristics underlying the theories and logics:
\begin{itemize}[noitemsep,nolistsep] %% these modifiers are not supported in all document classes
\item IA: integers
\item RA: reals
\item N: non-linear arithmetic
\item A or AX: arrays
\item UF: uninterpreted functions
\item QF\_: quantifier-free
\item BV: bit-vectors
\end{itemize}
This list of characteristics partially corresponds to the set of SMT-LIB theories. The exceptions are quantification and non-linearity, both of which designate lifting restrictions on the kinds of terms allowed in the resulting logic. (Since they do not correspond exactly to underlying SMT theories, the term {\em characteristics} is used in this discussion.)

It is possible to define a logic for any set of these characteristics in an almost purely combinatorial sense.
Barring non-linearity without arithmetic and the useless empty set, there are 111 combinations of characteristics (instead of the full 128). SMT-LIB has definitions and benchmarks for 23 of these logics. Fig.~\ref{Fig:lattice} shows what might be expected to be natural containment relationships among the logics defined at the time of this evaluation.

However, these containment relationships are not actually sub-logic relationships. That more accurate relationship is shown in a diagram on the SMT-LIB website: \url{http://smtlib.cs.uiowa.edu/logics.shtml} (due to Cesare Tinelli). The principal cause of this lack of sub-logic relationship is that many of the logics supporting arrays support only specific sorts of arrays. For example, the basic QF\_AX logic allows arbitrary new sort symbols, including arrays over those free sorts. However, QF\_ABV (arrays and bit-vectors) does not allow new sort symbols and only allows arrays over bit-vector sorts.

The naming convention for logics is almost but not quite simply a combination of the letters indicated in the list above,
corresponding to the characteristics of the logic, with the letters listed in order by convention: {\tt [QF\_][A|AX][UF][BV][N|L][IA|RA|IRA]}. The non-uniformities are these:
\begin{itemize}[noitemsep,nolistsep]
\item quantification is indicated by {\em removing} a {\tt QF\_} prefix;
\item the absence of non-linearity ({\tt N}), i.e., linearity, is indicated by an {\tt L}, instead of by no designator;
\item integer and real arithmetic are indicated by two letters ({\tt IA} and {\tt RA}) and their combination by {\tt IRA}; \item the {\tt A} used in the arithmetic designators could be ambiguous with the designator for arrays, except for position;
\item a logic with just arrays is named {\tt QF\_AX}, whereas in other combinations including arrays, just an {\tt A} instead of {\tt AX} is used.
\end{itemize}
For example, {\tt UFNIA} includes quantification (no {\tt QF\_} prefix), uninterpreted functions (UF), non-linearity (N), and integer arithmetic (IA), leaving out arrays, bit-vectors, and reals.

These names have roots in the history of development of SMT-LIB. Nevertheless as the set of logics grows---23 here and 34 by the 2014 competition---the situation will increasingly confuse new users of SMT-LIB trying to understand the structure of logics; in addition new theories and accompanying logics are anticipated.
It would be useful to standardize the naming convention along the lines of current use, perhaps with slight adjustments (e.g., renaming {\tt QF\_AX} to {\tt QF\_A} and removing the excess {\tt A} from {\tt IA} and {\tt RA}), in order to define precisely the logic implied by any combination of designators. A policy regarding names would provide a process for defining names for new theories and logics. This would
also make it easier to accurately characterize new benchmarks and to summarize the logics supported by a given solver. This need has been raised to the SMT-LIB coordinators and some work has been started to address it.

@@

\begin{figure}
\centering
\resizebox{\columnwidth}{!}{
%%\includegraphics[width=\textwidth]{LogicSupportBySolvers}
\input{LogicSupportBySolvers}
}

\caption{Support for each logic as stated by solver implementors. The dots instead of Xs identify the 2013 versions of solvers. The numbers in the bottom row are the number of 2013 solvers for each logic. In some cases, indicated by numbers in parentheses in the left-hand column, there are multiple versions of the same tool in different years with the same support for logics.}
\label{Fig:solversupport}
\end{figure}

@@

Note that the support indicated is that stated by the solver supplier. In some cases, a solver supports one logic but is not listed as supporting a subset of that logic. The general reason is that the subset, with restricted characteristics, allows for specific optimizations that are not implemented. The more general solver could act on problems of the
more restricted benchmarks but is not deemed competitive and thus was not formally entered into the competition for the restricted logic.

@@

Fig.~\ref{Fig:CumulativeTimes} shows the fraction of benchmarks completed (y-axis) within a given time (in seconds, on the x-axis); thus points toward the upper left are better (more completed in less time). The results for all logics are shown together. The four curves are the data for the four years from 2010 to 2013. The lowest (marked with fine dots) curve is that for 2010. There is clearly significant progress made from 2010 to later years: the number of benchmarks completed within a given time is noticeably higher. The curves for 2011 and 2012 (heavy dashes and short dashes, respectively) are clearly above 2010, but are not substantially different from each other. The solvers for 2013 (heavy solid line) are uniformly above those for previous years, but not by a large amount.

@@

\begin{figure}
\centering
\includegraphics[width=.8\textwidth]{CumulativeTimes3}
\caption{The fraction of benchmarks (y-axis) whose winning time is less than the given number of seconds (x-axis), by year.}
\label{Fig:CumulativeTimes}
\end{figure}

@@

\section*{Acknowledgments}
